---
title: "Secondary Structure Analysis"
author: "Sadhana Chidambaran"
output: html_document
---

```{r echo = FALSE}
knitr::opts_chunk$set(fig.align = "center", warning = FALSE, message = FALSE)
```

Load packages
```{r, message = FALSE}
library(ggpubr)
library(tidyverse)
```

### Create all functions

Function to reformat RNAfold results
```{r}
rnafold.reformat <- function(data) {
  
  data %>% 
    mutate(gene = gsub(">", "", gene)) %>%
    separate(fold_info, into = c("a", "b"), sep = " \\(", remove = FALSE) %>% 
    mutate(fold.energy = as.numeric(str_remove(b, "\\)"))) %>% 
    dplyr::rename("fold.structure" = "a") %>% 
    select(-c("fold_info", "b"))
  
}
```

Initialize vectors for all thresholds and bases
```{r}
thresholds <- c(0.005, 0.05, 0.1)
names(thresholds) <- thresholds
```

Function for Kolmogorov–Smirnov tests
```{r}
ks.test.func <- function(data) {
  
  # Iterate over all thresholds
  lapply(thresholds, function(x) {
    
    # Filter for sequences that start with a particular base, at a particular threshold for low fragmentation
    filtered.data <- data %>%
      filter(threshold == x)
    
    seq.data <- filtered.data %>% 
      filter(source == "seq")
    
    rand.data <- filtered.data %>% 
      filter(source == "rand")
    
    # KS test to determine if folding energy distribution is less in regions of low fragmentation vs. randomly selected sequences
    ks <- ks.test(seq.data$fold.energy, rand.data$fold.energy, alternative = "greater")
    
    # Round p-values to 5 decimal places
    ks$p.value %>% 
      format(digits = 5, nsmall = 5)
    }) %>% 
    unlist() %>% 
    as.data.frame() %>% 
    rownames_to_column() %>% 
    dplyr::rename("threshold" = "rowname", "p.value" = ".")
}
```

### Run RNAfold on all the data for the 100 most expressed genes

Run RNAfold on all sequences (those from regions of low fragmentation and those that were randomly selected)
```{bash}
# Run RNAfold
for file in /data/sadhana/rnaseq_bias/fastas/100_genes/*.fasta; do

  name=`echo $file | cut -d '/' -f 7`
  path=/data/sadhana/rnaseq_bias/secondary_structure_analysis/100_genes/output/${name%.*}
  
  RNAfold --noPS $file > $path\.out
  
done

# Convert RNAfold output to CSVs
for file in /data/sadhana/rnaseq_bias/secondary_structure_analysis/100_genes/output/*.out; do

  name=`echo $file | cut -d '/' -f 8`
  path=/data/sadhana/rnaseq_bias/secondary_structure_analysis/100_genes/csvs/${name%.*}
  
  echo "gene,sequence,fold_info" > $path.csv
  while read one; read two; read three
  do
    echo "$one","$two","$three"
  done < $file >> $path.csv
  
done
```

### Get RNAfold results for the 100 most expressed genes

Read in RNAfold results
```{r, message = FALSE}
# Get all output files
rnafold.files.100.edges <- grep("edge", 
                                dir("/data/sadhana/rnaseq_bias/secondary_structure_analysis/100_genes/csvs", 
                                    full.names = TRUE), 
                                value = TRUE)
names(rnafold.files.100.edges) <- lapply(rnafold.files.100.edges, function(filename) {
  filename <- str_split(filename, "/")[[1]][8] %>% 
    gsub(".csv", "", .)
  str_split(filename, "_")[[1]][5]
}) %>% 
  unlist()

rnafold.files.100.seqs <- setdiff(dir("/data/sadhana/rnaseq_bias/secondary_structure_analysis/100_genes/csvs", 
                                      full.names = TRUE), 
                                  rnafold.files.100.edges)
names(rnafold.files.100.seqs) <- lapply(rnafold.files.100.seqs, function(filename) {
  filename <- str_split(filename, "/")[[1]][8] %>% 
    gsub(".csv", "", .)
  str_split(filename, "_")[[1]][4]
}) %>% 
  unlist()

# Read in all output files
rnafold.100.seqs.5.end <- lapply(grep("seqs_5_ends", rnafold.files.100.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.100.seqs.3.end <- lapply(grep("seqs_3_ends", rnafold.files.100.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.100.rand.5.end <- lapply(grep("rand_5_ends", rnafold.files.100.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.100.rand.3.end <- lapply(grep("rand_3_ends", rnafold.files.100.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })

rnafold.100.seqs.5.edge <- lapply(grep("seqs_5_ends", rnafold.files.100.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.100.seqs.3.edge <- lapply(grep("seqs_3_ends", rnafold.files.100.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.100.rand.5.edge <- lapply(grep("rand_5_ends", rnafold.files.100.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.100.rand.3.edge <- lapply(grep("rand_3_ends", rnafold.files.100.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
```

Reformat RNAfold results
```{r}
rnafold.100.seqs.5.end <- lapply(rnafold.100.seqs.5.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.100.seqs.3.end <- lapply(rnafold.100.seqs.3.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.100.rand.5.end <- lapply(rnafold.100.rand.5.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.100.rand.3.end <- lapply(rnafold.100.rand.3.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.100.seqs.5.edge <- lapply(rnafold.100.seqs.5.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.100.seqs.3.edge <- lapply(rnafold.100.seqs.3.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.100.rand.5.edge <- lapply(rnafold.100.rand.5.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.100.rand.3.edge <- lapply(rnafold.100.rand.3.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")
```

### Generate box plots for folding energy distributions

```{r}
# Combine RNAfold results for 5' read ends
rnafold.100.5.data <- rbind(rnafold.100.seqs.5.end, rnafold.100.rand.5.end) %>% 
  mutate(seq.start = substr(sequence, 1, 1)) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.100.5.data$source <- factor(rnafold.100.5.data$source , levels = c("seq", "rand"))

# Combine RNAfold results for 3' read ends
rnafold.100.3.data <- rbind(rnafold.100.seqs.3.end, rnafold.100.rand.3.end) %>% 
  mutate(seq.end = substr(sequence, nchar(sequence), nchar(sequence))) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.100.3.data$source <- factor(rnafold.100.3.data$source , levels = c("seq", "rand"))

# Combine RNAfold results for 5' edges
rnafold.100.5.data.edges <- rbind(rnafold.100.seqs.5.edge, rnafold.100.rand.5.edge) %>% 
  mutate(seq.start = substr(sequence, 1, 1)) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.100.5.data.edges$source <- factor(rnafold.100.5.data.edges$source , levels = c("seq", "rand"))

# Combine RNAfold results for 3' edges
rnafold.100.3.data.edges <- rbind(rnafold.100.seqs.3.edge, rnafold.100.rand.3.edge) %>% 
  mutate(seq.end = substr(sequence, nchar(sequence), nchar(sequence))) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.100.3.data.edges$source <- factor(rnafold.100.3.data.edges$source , levels = c("seq", "rand"))
```

Generate box plots for folding energy distributions (5' read ends)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.100.5.ks <- ks.test.func(rnafold.100.5.data)

plot.100.genes.5 <- ggplot(rnafold.100.5.data, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "100 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("Blocks", "Random sequences")) + 
  geom_text(data = rnafold.100.5.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_100_5_ends.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (3' read ends)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.100.3.ks <- ks.test.func(rnafold.100.3.data)

plot.100.genes.3 <- ggplot(rnafold.100.3.data, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "100 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("Blocks", "Random sequences")) + 
  geom_text(data = rnafold.100.3.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_100_3_ends.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (5' edges)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.100.5.edges.ks <- ks.test.func(rnafold.100.5.data.edges)

plot.100.genes.5.edges <- ggplot(rnafold.100.5.data.edges, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "100 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("5' Edges", "Random sequences")) + 
  geom_text(data = rnafold.100.5.edges.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_100_5_edges.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (3' edges)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.100.3.edges.ks <- ks.test.func(rnafold.100.3.data.edges)

plot.100.genes.3.edges <- ggplot(rnafold.100.3.data.edges, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "100 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("3' Edges", "Random sequences")) + 
  geom_text(data = rnafold.100.3.edges.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_100_3_edges.png", width = 9, height = 5)
```

### Run RNAfold on all the data for the 200 most expressed genes

Run RNAfold on all sequences (those from regions of low fragmentation and those that were randomly selected)
```{bash}
# Run RNAfold
for file in /data/sadhana/rnaseq_bias/fastas/200_genes/*.fasta; do

  name=`echo $file | cut -d '/' -f 7`
  path=/data/sadhana/rnaseq_bias/secondary_structure_analysis/200_genes/output/${name%.*}
  
  RNAfold --noPS $file > $path\.out
  
done

# Convert RNAfold output to CSVs
for file in /data/sadhana/rnaseq_bias/secondary_structure_analysis/200_genes/output/*.out; do

  name=`echo $file | cut -d '/' -f 8`
  path=/data/sadhana/rnaseq_bias/secondary_structure_analysis/200_genes/csvs/${name%.*}
  
  echo "gene,sequence,fold_info" > $path.csv
  while read one; read two; read three
  do
    echo "$one","$two","$three"
  done < $file >> $path.csv
  
done
```

### Get RNAfold results for the 200 most expressed genes

Read in RNAfold results
```{r, message = FALSE}
# Get all output files
rnafold.files.200.edges <- grep("edge", 
                                dir("/data/sadhana/rnaseq_bias/secondary_structure_analysis/200_genes/csvs", 
                                    full.names = TRUE), 
                                value = TRUE)
names(rnafold.files.200.edges) <- lapply(rnafold.files.200.edges, function(filename) {
  filename <- str_split(filename, "/")[[1]][8] %>% 
    gsub(".csv", "", .)
  str_split(filename, "_")[[1]][5]
}) %>% 
  unlist()

rnafold.files.200.seqs <- setdiff(dir("/data/sadhana/rnaseq_bias/secondary_structure_analysis/200_genes/csvs", 
                                      full.names = TRUE), 
                                  rnafold.files.200.edges)
names(rnafold.files.200.seqs) <- lapply(rnafold.files.200.seqs, function(filename) {
  filename <- str_split(filename, "/")[[1]][8] %>% 
    gsub(".csv", "", .)
  str_split(filename, "_")[[1]][4]
}) %>% 
  unlist()

# Read in all output files
rnafold.200.seqs.5.end <- lapply(grep("seqs_5_ends", rnafold.files.200.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.200.seqs.3.end <- lapply(grep("seqs_3_ends", rnafold.files.200.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.200.rand.5.end <- lapply(grep("rand_5_ends", rnafold.files.200.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.200.rand.3.end <- lapply(grep("rand_3_ends", rnafold.files.200.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })

rnafold.200.seqs.5.edge <- lapply(grep("seqs_5_ends", rnafold.files.200.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.200.seqs.3.edge <- lapply(grep("seqs_3_ends", rnafold.files.200.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.200.rand.5.edge <- lapply(grep("rand_5_ends", rnafold.files.200.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.200.rand.3.edge <- lapply(grep("rand_3_ends", rnafold.files.200.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
```

Reformat RNAfold results
```{r}
rnafold.200.seqs.5.end <- lapply(rnafold.200.seqs.5.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.200.seqs.3.end <- lapply(rnafold.200.seqs.3.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.200.rand.5.end <- lapply(rnafold.200.rand.5.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.200.rand.3.end <- lapply(rnafold.200.rand.3.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.200.seqs.5.edge <- lapply(rnafold.200.seqs.5.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.200.seqs.3.edge <- lapply(rnafold.200.seqs.3.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.200.rand.5.edge <- lapply(rnafold.200.rand.5.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.200.rand.3.edge <- lapply(rnafold.200.rand.3.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")
```

### Generate box plots for folding energy distributions

```{r}
# Combine RNAfold results for 5' read ends
rnafold.200.5.data <- rbind(rnafold.200.seqs.5.end, rnafold.200.rand.5.end) %>% 
  mutate(seq.start = substr(sequence, 1, 1)) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.200.5.data$source <- factor(rnafold.200.5.data$source , levels = c("seq", "rand"))

# Combine RNAfold results for 3' read ends
rnafold.200.3.data <- rbind(rnafold.200.seqs.3.end, rnafold.200.rand.3.end) %>% 
  mutate(seq.end = substr(sequence, nchar(sequence), nchar(sequence))) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.200.3.data$source <- factor(rnafold.200.3.data$source , levels = c("seq", "rand"))

# Combine RNAfold results for 5' edges
rnafold.200.5.data.edges <- rbind(rnafold.200.seqs.5.edge, rnafold.200.rand.5.edge) %>% 
  mutate(seq.start = substr(sequence, 1, 1)) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.200.5.data.edges$source <- factor(rnafold.200.5.data.edges$source , levels = c("seq", "rand"))

# Combine RNAfold results for 3' edges
rnafold.200.3.data.edges <- rbind(rnafold.200.seqs.3.edge, rnafold.200.rand.3.edge) %>% 
  mutate(seq.end = substr(sequence, nchar(sequence), nchar(sequence))) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.200.3.data.edges$source <- factor(rnafold.200.3.data.edges$source , levels = c("seq", "rand"))
```

Generate box plots for folding energy distributions (5' read ends)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.200.5.ks <- ks.test.func(rnafold.200.5.data)

plot.200.genes.5 <- ggplot(rnafold.200.5.data, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "200 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("Blocks", "Random sequences")) + 
  geom_text(data = rnafold.200.5.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_200_5_ends.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (3' read ends)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.200.3.ks <- ks.test.func(rnafold.200.3.data)

plot.200.genes.3 <- ggplot(rnafold.200.3.data, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "200 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("Blocks", "Random sequences")) + 
  geom_text(data = rnafold.200.3.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_200_3_ends.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (5' edges)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.200.5.edges.ks <- ks.test.func(rnafold.200.5.data.edges)

plot.200.genes.5.edges <- ggplot(rnafold.200.5.data.edges, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "200 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("5' Edges", "Random sequences")) + 
  geom_text(data = rnafold.200.5.edges.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_200_5_edges.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (3' edges)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.200.3.edges.ks <- ks.test.func(rnafold.200.3.data.edges)

plot.200.genes.3.edges <- ggplot(rnafold.200.3.data.edges, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "200 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("3' Edges", "Random sequences")) + 
  geom_text(data = rnafold.200.3.edges.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_200_3_edges.png", width = 9, height = 5)
```

### Run RNAfold on all the data for the 500 most expressed genes

Run RNAfold on all sequences (those from regions of low fragmentation and those that were randomly selected)
```{bash}
# Run RNAfold
for file in /data/sadhana/rnaseq_bias/fastas/500_genes/*.fasta; do

  name=`echo $file | cut -d '/' -f 7`
  path=/data/sadhana/rnaseq_bias/secondary_structure_analysis/500_genes/output/${name%.*}
  
  RNAfold --noPS $file > $path\.out
  
done

# Convert RNAfold output to CSVs
for file in /data/sadhana/rnaseq_bias/secondary_structure_analysis/500_genes/output/*.out; do

  name=`echo $file | cut -d '/' -f 8`
  path=/data/sadhana/rnaseq_bias/secondary_structure_analysis/500_genes/csvs/${name%.*}
  
  echo "gene,sequence,fold_info" > $path.csv
  while read one; read two; read three
  do
    echo "$one","$two","$three"
  done < $file >> $path.csv
  
done
```

### Get RNAfold results for the 500 most expressed genes

Read in RNAfold results
```{r, message = FALSE}
# Get all output files
rnafold.files.500.edges <- grep("edge", 
                                dir("/data/sadhana/rnaseq_bias/secondary_structure_analysis/500_genes/csvs", 
                                    full.names = TRUE), 
                                value = TRUE)
names(rnafold.files.500.edges) <- lapply(rnafold.files.500.edges, function(filename) {
  filename <- str_split(filename, "/")[[1]][8] %>% 
    gsub(".csv", "", .)
  str_split(filename, "_")[[1]][5]
}) %>% 
  unlist()

rnafold.files.500.seqs <- setdiff(dir("/data/sadhana/rnaseq_bias/secondary_structure_analysis/500_genes/csvs", 
                                      full.names = TRUE), 
                                  rnafold.files.500.edges)
names(rnafold.files.500.seqs) <- lapply(rnafold.files.500.seqs, function(filename) {
  filename <- str_split(filename, "/")[[1]][8] %>% 
    gsub(".csv", "", .)
  str_split(filename, "_")[[1]][4]
}) %>% 
  unlist()

# Read in all output files
rnafold.500.seqs.5.end <- lapply(grep("seqs_5_ends", rnafold.files.500.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.500.seqs.3.end <- lapply(grep("seqs_3_ends", rnafold.files.500.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.500.rand.5.end <- lapply(grep("rand_5_ends", rnafold.files.500.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.500.rand.3.end <- lapply(grep("rand_3_ends", rnafold.files.500.seqs, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })

rnafold.500.seqs.5.edge <- lapply(grep("seqs_5_ends", rnafold.files.500.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.500.seqs.3.edge <- lapply(grep("seqs_3_ends", rnafold.files.500.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.500.rand.5.edge <- lapply(grep("rand_5_ends", rnafold.files.500.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
rnafold.500.rand.3.edge <- lapply(grep("rand_3_ends", rnafold.files.500.edges, value = TRUE), 
                             function(x) { read_csv(x, col_names = TRUE) })
```

Reformat RNAfold results
```{r}
rnafold.500.seqs.5.end <- lapply(rnafold.500.seqs.5.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.500.seqs.3.end <- lapply(rnafold.500.seqs.3.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.500.rand.5.end <- lapply(rnafold.500.rand.5.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.500.rand.3.end <- lapply(rnafold.500.rand.3.end, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.500.seqs.5.edge <- lapply(rnafold.500.seqs.5.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.500.seqs.3.edge <- lapply(rnafold.500.seqs.3.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "seq")

rnafold.500.rand.5.edge <- lapply(rnafold.500.rand.5.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")

rnafold.500.rand.3.edge <- lapply(rnafold.500.rand.3.edge, rnafold.reformat) %>% 
  bind_rows(.id = "threshold") %>% 
  mutate(source = "rand")
```

### Generate box plots for folding energy distributions

```{r}
# Combine RNAfold results for 5' read ends
rnafold.500.5.data <- rbind(rnafold.500.seqs.5.end, rnafold.500.rand.5.end) %>% 
  mutate(seq.start = substr(sequence, 1, 1)) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.500.5.data$source <- factor(rnafold.500.5.data$source , levels = c("seq", "rand"))

# Combine RNAfold results for 3' read ends
rnafold.500.3.data <- rbind(rnafold.500.seqs.3.end, rnafold.500.rand.3.end) %>% 
  mutate(seq.end = substr(sequence, nchar(sequence), nchar(sequence))) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.500.3.data$source <- factor(rnafold.500.3.data$source , levels = c("seq", "rand"))

# Combine RNAfold results for 5' edges
rnafold.500.5.data.edges <- rbind(rnafold.500.seqs.5.edge, rnafold.500.rand.5.edge) %>% 
  mutate(seq.start = substr(sequence, 1, 1)) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.500.5.data.edges$source <- factor(rnafold.500.5.data.edges$source , levels = c("seq", "rand"))

# Combine RNAfold results for 3' edges
rnafold.500.3.data.edges <- rbind(rnafold.500.seqs.3.edge, rnafold.500.rand.3.edge) %>% 
  mutate(seq.end = substr(sequence, nchar(sequence), nchar(sequence))) %>% 
  filter(threshold %in% c("0.005", "0.05", "0.1"))
rnafold.500.3.data.edges$source <- factor(rnafold.500.3.data.edges$source , levels = c("seq", "rand"))
```

Generate box plots for folding energy distributions (5' read ends)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.500.5.ks <- ks.test.func(rnafold.500.5.data)

plot.500.genes.5 <- ggplot(rnafold.500.5.data, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "500 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("Blocks", "Random sequences")) + 
  geom_text(data = rnafold.500.5.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_500_5_ends.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (3' read ends)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.500.3.ks <- ks.test.func(rnafold.500.3.data)

plot.500.genes.3 <- ggplot(rnafold.500.3.data, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "500 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("Blocks", "Random sequences")) + 
  geom_text(data = rnafold.500.3.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_500_3_ends.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (5' edges)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.500.5.edges.ks <- ks.test.func(rnafold.500.5.data.edges)

plot.500.genes.5.edges <- ggplot(rnafold.500.5.data.edges, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "500 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("5' Edges", "Random sequences")) + 
  geom_text(data = rnafold.500.5.edges.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_500_5_edges.png", width = 9, height = 5)
```

Generate box plots for folding energy distributions (3' edges)
```{r, warning = FALSE, fig.width = 9, fig.height = 5}
# Conduct KS tests between folding energy distributions of regions of low fragmentation and randomly selected sequences
rnafold.500.3.edges.ks <- ks.test.func(rnafold.500.3.data.edges)

plot.500.genes.3.edges <- ggplot(rnafold.500.3.data.edges, aes(x = threshold, y = fold.energy)) + 
  geom_boxplot(aes(fill = source)) + 
  labs(title = "500 most expressed genes", x = "Threshold", y = "Free energy (kcal/mol)") + 
  theme(legend.title = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12), 
        text = element_text(size = 12), axis.text = element_text(size = 12)) + 
  scale_fill_discrete(labels = c("3' Edges", "Random sequences")) + 
  geom_text(data = rnafold.500.3.edges.ks, mapping = aes(x = threshold, y = 0.5, label = p.value), inherit.aes = FALSE, size = 4) + 
  ylim(-13, 1)

ggsave("rnafold_500_3_edges.png", width = 9, height = 5)
```

### Arrange box plots

```{r}
ggarrange(plot.100.genes.5, plot.200.genes.5, plot.500.genes.5, 
          nrow = 1, ncol = 3, common.legend = TRUE, legend = "bottom")

ggsave("rnafold_blocks.png", width = 12, height = 5, dpi = "retina")

ggarrange(plot.100.genes.5.edges, plot.200.genes.5.edges, plot.500.genes.5.edges, 
          nrow = 1, ncol = 3, common.legend = TRUE, legend = "bottom")

ggsave("rnafold_5_edges.png", width = 12, height = 5, dpi = "retina")

ggarrange(plot.100.genes.3.edges, plot.200.genes.3.edges, plot.500.genes.3.edges, 
          nrow = 1, ncol = 3, common.legend = TRUE, legend = "bottom")

ggsave("rnafold_3_edges.png", width = 12, height = 5, dpi = "retina")
```
